---
title: "Mid-project update"
author: "Rachel Rodriguez, Riho Ishida, and Alicia Magsano"
date: "2024-11-19"
output: pdf_document
---

```{r}
library(tidyverse)
```

# Previous Work/Information

*Batrachochytrium dendrobatidis* (*Bd*) is a fungal pathogen that causes chytridiomycosis which is a disease that is lethal to a large number of amphibian species (Scheele et al., 2019). Chytridiomycosis has had detrimental effects on amphibian species as it has caused global declines and extinctions in many susceptible species (Scheele et al., 2019; Skerratt et al., 2007). As a result, the declines caused by *Bd* have been called, "the greatest recorded loss of biodiversity attributable to a disease" (Scheele et al., 2019). 

Given how severe the effects of *Bd* are it is important to understand what factors can help to limit the effects of this deadly pathogen. Temperature may be one such factor as previous research has shown that *Bd*, in its stationary phase, thrives in colder temperatures of 13-15°C, with a rapid decline of density at around 26°C (Stevenson et al., 2013). The same study demonstrated that the parasite grows best at around 21°C (Stevenson et al., 2013). Furthermore, there is evidence that *Bd* zoospore activity decreases as temperature increases, which correlates with a reduced prevalence of *Bd* in hosts (Sapsford et al., 2013; Stevenson et al., 2013). In vivo experiments support this, as reduced mortality is seen in subjects receiving higher temperatures for treatment (Berger et al., 2004). Therefore, given that *Bd* favours cooler temperatures, it is important to understand how changes in temperature due to climate change may affect *Bd* because this can help us better understand how to facilitate future conservation efforts for amphibians. 

Other studies have used modelling to determine the distribution of *Bd* in a certain location or to predict how *Bd*'s range may shift as a result of climate change (Sun et al., 2023; Tytar et al., 2023). However, there are not any studies on how the change in temperature due to climate change may change the dynamics of *Bd* and how it spreads within a population. 

# Our Question and Prediction

Question: How will changes in temperature due to climate change affect the spread of *Bd* within a population of frogs? 
- Specifically, how will the number of zoospores and the number of infected/susceptible individuals change over time given there is a change in the environmental temperature? 

Prediction: Since *Bd* favours cooler temperatures, we predict that over time at higher temperatures there will be fewer zoospores and thus fewer infected individuals within the population. 

# Analysis Plan

We planned to create a thermal performance curve (TPC) from data containing zoospore count and temperature to answer this question. Then we would make a stage-structured SIR model for a frog population which we would solve for at different temperatures by using the TPC we created to indicate the initial number of zoospores present at a given temperature. 

## Data Used

The data we used to make the TPC was created by combining the data for three different studies including two field studies and one lab study.

1. First Field Study (Kriger & Hero, 2007)
We used the data collected by Kriger and Hero (2007) from their study on the seasonal variation of chytridiomycosis prevalence and severity in a single population of *Litoria wilcoxii* (eastern stony creek frogs). In this study, they captured and sampled *L. wilcoxii* over 21 months in 6-week intervals within a 1 km area of the Nerang River in Queensland, Australia. During each of the 13 sampling sessions, they swabbed frogs for chytridiomycosis, recorded the air and water temperature at the site, and obtained rainfall data from a nearby weather station. The swabs were tested for *Bd* presence using quantitative polymerase chain reaction (qPCR) techniques which were used to determine the prevalence of *Bd* (divided the number of *Bd* positive frogs by the total number of frogs sampled that session) and the geometric mean number of zoospores (parasite load) found in infected frogs. Additionally, they determined each session's air and water temperature by calculating the mean air temperature and total amount of rain in the 30 days leading up to the sampling day. Finally, since the water temperature was manually sampled at the beginning and end of each sampling session, the mean of these two measurements was used as the water temperature for a given sampling session. Therefore, we used the zoospores and 30-day air temperature (°C) data from this study. 

2. Second Field Study (Kriger et al., 2007)
We also used the data collected by Kriger et al. (2007) whose study focused on how the prevalence and intensity of *Bd* is affected by latitudinal variation. In this study, they captured and swabbed *Litoria lesueuri* (stony creek frogs) for *Bd* infection at 31 sites along the east coast of Australia which encompassed a gradient of latitudes and the sampling from each site was done within a 42-day period (20 September 2005 – 1 November 2005). The swabs collected at each site were tested for *Bd* presence and quantity using qPCR. Furthermore, in addition to recording the latitude, longitude, and elevation at each sampling site, Kriger et al. (2007) also recorded different short and long-term climatic conditions. From the climatic conditions recorded, we were interested in the mean temperature of the warmest quarter (MTWQ). Therefore, we used the zoospores per frog (mean number of zoospores of all frogs sampled at a site) and MTWQ (°C) data from this study. 

3. Lab Study (Sheets et al., 2021)
Additionally, we used the data collected by Sheets et al. (2021) who conducted a study to determine how temperature affects different *Bd* isolates. In this study, Sheets et al. (2021) specifically tested isolates from the *Bd* lineage Global Panzootic Lineage (*Bd*GPL) which is hypervirulent (Farrer et al., 2011). In their study, Sheets et al. (2021) collected 5 different *Bd* isolates from amphibians around the United States which they genotyped using an amplicon sequencing approach. They then tested how temperature affected different traits of each isolate including viability, zoospore density, growth rate, and carrying capacity by incubating the isolates at various temperatures (4, 12, 17, 21, 25, 26, and 27 °C). Using this information, they generated thermal performance curves for each genotype and isolate. Therefore, we used the temperature and zoospore count (calculated as the average number of zoospores counted from two sides of a hemocytometer) data from the Temp and Counts columns. However, since the zoospore count was recorded over multiple days for each isolate and temperature combination we only included the highest number of zoospores for each isolate and temperature combination in our combined dataset. We decided to do this because before the highest number of zoospores was reached the number of zoospores was still able to grow and after the highest number was reached the number of zoospores likely began to decline due to not being able to persist in the lab setting anymore. 

## TPC

```{r}
data <- read.csv("Combined Chytrid Data.csv") 

# Max Temperature in the combined dataset
max(data$Temperature)
data$Zoospores <- as.integer(data$Zoospores)

# Plot of the combined zoospore and temperature data
plot(data$Temperature,data$Zoospores, 
     main = "Change in Zoospore Count Over Different Temperatures", 
     xlab = "Temperature (°C)",
     ylab = "Number of Zoospores")

# Distribution of zoospore count in the combined dataset 
hist(data$Zoospores, main = "Distribution of Zoospore Count", 
     xlab = "Number of Zoospores")
```

Since our data had so much variance we used a negative binomial distribution to construct the TPC. 

```{r}

```


## Mathematical Model



## Assumptions and Limitations



## Remaining Plan 



# References 

Berger, L., Speare, R., Hines, H., Martelli, G., Hyatt, A., McDonald, K., Skerratt, L., Olsen, 
  V., Clarke, J., Gillespie, G., Mahony, M., Sheppard, N., Williams, C., & Tyler, M. (2004). 
  Effect Of Season And Temperature On Mortality In Amphibians Due To Chytridiomycosis. 
  *Australian Veterinary Journal, 82*(7), 434–439. 
  https://doi.org/10.1111/j.1751-0813.2004.tb11137.x

Kriger, K. M., & Hero, J. M. (2007). Large-scale seasonal variation in the prevalence and 
  severity of chytridiomycosis. *Journal of Zoology (1987), 271*(3), 352–359. 
  https://doi.org/10.1111/j.1469-7998.2006.00220.x

Kriger, K. M., Pereoglou, F., & Hero, J. M. (2007). Latitudinal Variation in the Prevalence and 
  Intensity of Chytrid (Batrachochytrium dendrobatidis) Infection in Eastern Australia. 
  *Conservation Biology, 21*(5), 1280–1290. https://doi.org/10.1111/j.1523-1739.2007.00777.x

Sapsford, S. J., Alford, R. A., & Schwarzkopf, L. (2013). Elevation, Temperature, and Aquatic 
  Connectivity All Influence the Infection Dynamics of the Amphibian Chytrid Fungus in Adult 
  Frogs. *PloS One, 8*(12), e82425–e82425. https://doi.org/10.1371/journal.pone.0082425

Scheele, B. C., Pasmans, F., Skerratt, L. F., Berger, L., Martel, A., Beukema, W., Acevedo, A. 
  A., Burrowes, P. A., Carvalho, T., Catenazzi, A., De la Riva, I., Fisher, M. C., Flechas, S. 
  V., Foster, C. N., Frías-Álvarez, P., Garner, T. W. J., Gratwicke, B., Guayasamin, J. M.,   
  Hirschfeld, M., … Canessa, S. (2019). Amphibian fungal panzootic causes catastrophic and    
  ongoing loss of biodiversity. *Science, 363*(6434), 1459–1463. 
  https://doi.org/10.1126/science.aav0379

Sheets, C. N., Schmidt, D. R., Hurtado, P. J., Byrne, A. Q., Rosenblum, E. B., 
  Richards-Zawacki, C. L., & Voyles, J. (2021). Thermal Performance Curves of Multiple Isolates
  of Batrachochytrium dendrobatidis, a Lethal Pathogen of Amphibians. *Frontiers in Veterinary*
  *Science, 8*, 687084–687084. https://doi.org/10.3389/fvets.2021.687084

Skerratt, L. F., Berger, L., Speare, R., Cashins, S., McDonald, K. R., Phillott, A. D., Hines, 
  H. B., & Kenyon, N. (2007). Spread of Chytridiomycosis Has Caused the Rapid Global Decline 
  and Extinction of Frogs. *EcoHealth, 4*(2), 125–134. 
  https://doi.org/10.1007/s10393-007-0093-5

Stevenson, L. A., Alford, R. A., Bell, S. C., Roznik, E. A., Berger, L., & Pike, D. A. (2013). 
  Variation in Thermal Performance of a Widespread Pathogen, the Amphibian Chytrid Fungus 
  *Batrachochytrium dendrobatidis*. *PLoS ONE, 8*(9), e73830. 
  https://doi.org/10.1371/journal.pone.0073830

Sun, D., Ellepola, G., Herath, J., & Meegaskumbura, M. (2023). The two chytrid pathogens of 
  amphibians in Eurasia—climatic niches and future expansion. *BMC Ecology and Evolution,* 
  *23*(1), 26–26. https://doi.org/10.1186/s12862-023-02132-y

Tytar, V., Nekrasova, O., Pupins, M., Skute, A., Kirjušina, M., Gravele, E., Mezaraupe, L., 
  Marushchak, O., Čeirāns, A., Kozynenko, I., & Kulikova, A. A. (2023). Modeling the 
  Distribution of the Chytrid Fungus Batrachochytrium dendrobatidis with Special Reference to 
  Ukraine. *Journal of Fungi, 9*(6), 607-. https://doi.org/10.3390/jof9060607







